Viscoelastic Kelvin-Helmholtz instability

### 2015-07-03 paper/paper.tex
D_domain.py : added cline support

%	paper.tex	Toby Searle
%	Last modified: Wed 14 May 22:10:55 2014
%	using jfm template

@@

It was long thought that the elasticity of a viscoelastic fluid only served to
dampen instability. This view stems from early results showing that drag ina
turbulent Newtonian flow can be reduced by the addition of small amounts of
viscoelastic fluid (\cite{Toms1977}). Although drag reduction has recieved a
lot of attention in the past the studies above give reason to believe that
there is a potentially more prevalent and useful phenomenon at work in
viscoelastic fluids. It seems that viscoelastic turbulence is present at very
low Reynolds number, where it is driven by the elasticity of the polymeric
fluid rather than its inertia.

In 1997 Fabian Waleffe identified a Newtonian self sustaining process in plane
Couette flow \cite{Waleffe1997}. Streamwise rolls redistribute the streamwise
velocity into a streaky flow. This streaky flow is unstable through a
Kelvin-Helmholtz instability leading to a symmetry bifurcation to a three
dimensional perturbation which in turn re-energises the original streamwise
rolls. This exact solution to the Navier-Stokes equations is thought to be a
component of the transition to turbulence. In 1998 Waleffe constructed a
bifurcation diagram for this exact solution \cite{Waleffe1998} containing what
looks like a bifurcation from infinity. This is the behaviour expected of the
transition to turbulence in Couette flow.

@@

The key dimensionless number for turbulence without inertia in viscoelastic
flows is the Weissenberg number, which is the ratio of the normal stress to the
shear stress of the polymer component of the fluid $Wi = \frac{N_{1}}{\sigma}$.
For small shear rates, this is approximately equivalent to $De$, the Deborah
number, so that $Wi \approx De = \lambda \dot{\gamma}$ where $\dot{\gamma}$ is
the shear rate.

@@

$U$ and $V$ are the base streamwise (x) and wall normal (y) velocities
respectively. $T$ is the base stress tensor. We use Gauss-Labatto points in the
wall-normal ($y$) direction for the base profile, and decomposed the
disturbances to this flow into Fourier modes.  
\begin{equation}
    g(y) = \sum\limits_{n=-N}^{N} \widetilde{g}(y) e^{ikx + \lambda t}
\end{equation}
for all disturbance variables $g = u, v, p, \tau_{i,j}$. Where $N$ is the
number of Fourier modes, $k$ is the streamwise wavenumber of the disturbance
and $\lambda$ is the growth rate of the mode. This code uses two domains of
pseudo-spectral points to give increased resolution in the centre of the
simulation. The simulation was also repeated using a code for which the system
of points was stretched using: 
\begin{align}
    z_{u} &= \frac{y_{u}}{1-y_{u}} \\
    z_{b} &= \frac{y_{b}}{1+y_{b}} \\
    U     &= \tanh{z/\Delta} 
    \label{eq:KH_inf_profile}
\end{align}
This system ought to be completely insensitive to the boundary conditions on
the flow and so also behave as a free shear instability.

@@

Dependence of the instability on the channel walls persists at very small
$\Delta$. We believe that this is due to the inertial effects at low Reynold's
number being so weak. This means it is very easy to move the fluid even in at
the shear rate present far from the centre of the channel. I think the
instability generated by the elasticity of the fluid needs the walls to dampen
it for some reason. The fluid moves more as one block because it has high a
viscosity. This one block is easier to get going because of the low inertia it
is also easy to stop. It then needs something solid to kill the fluid motion,
this is either a very large body of stationary fluid or a wall. The inertial
effects must reduce the velocity of the fluid far from the walls so that it can
feel the effect of the streamwise component of the velocity gradient.

@@

\begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{KH_onset_beta_Wi}
    \caption{}
    \label{fig:KH_elastic_onset}
\end{subfigure}
\caption{
a) The change in the height of the dispersion relation as $\Wi$ is increased at
$\Rey = 0$ with $\Delta = 0.1$. This instability must be entirely elastic in
nature. b) The change in the height of the dispersion relation as the Reynolds
number is reduced at $\Wi=5$. At low $\Rey$ the instability is clearly still
present for large polymer concentrations, consistent with purely elastic
turbulence with $\Delta = 0.1$.  c) The value of the Weissenberg number at
which the velocity profile becomes unstable against $\beta$ at $\Delta = 0.1$.
The onset of the purely elastic instability moves to higher $Wi$ as $\beta$
increases.
}
\end{figure}

@@

In conclusion, we can say that there is a Kelvin-Helmholtz like instability for
purely elastic shear flows. The unstable region is much larger than that seen
in Newtonian fluids relative to the region of shear. The wavenumber of the
instability is much smaller in the viscoelastic case ($k<0.06$), suggesting a
larger minimal flow unit is required in viscoelastic fluids. This instability
is enhanced by introducing a finite extensibility to the polymers. The
instability has grown to its maximum by the time it reaches an effective
Weissenberg number $ Wi \sim 50$ after which it is saturated. The instability
is also strongly dependent on the walls at low effective Reynold's number.

>>>>

%	paper.tex	Toby Searle
%	Last modified: Fri  6 Jun 15:39:17 2014
%	using jfm template

@@

It was long thought that the elasticity of a viscoelastic fluid only served to
dampen instability. This view stems from early results showing that drag in a
turbulent Newtonian flow can be reduced by the addition of small amounts of
viscoelastic fluid (\cite{Toms1977}). Although drag reduction has received a
lot of attention in the past, studies on turbulence without inertia give reason
to believe that there may be purely elastic instabilities present in elastic
fluids relevent to industry.  This purely elastic turbulence is present at very
low Reynolds number, where it is driven by the elasticity of the polymeric
fluid rather than its inertia.

In 1997 Fabian Waleffe identified a Newtonian self sustaining process in plane
Couette flow \cite{Waleffe1997}. Streamwise rolls redistribute the streamwise
velocity into a streaky flow. This streaky flow is unstable through a
Kelvin-Helmholtz instability leading to a symmetry bifurcation to a three
dimensional flow. Finally, the nonlinear effects due to this instability
re-energises the original streamwise rolls. This exact solution to the
Navier-Stokes equations is thought to be a component of the transition to
turbulence. In 1998 Waleffe constructed a bifurcation diagram for this exact
solution \cite{Waleffe1998} containing what looks like a bifurcation from
infinity, just the behaviour expected of the transition to turbulence in plane
Couette flow.

@@

The key dimensionless number for turbulence without inertia in viscoelastic
flows is the Weissenberg number, which is the ratio of the normal stress to the
shear stress of the polymer component of the fluid $Wi = \frac{N_{1}}{\sigma} =
\lambda \dot{\gamma}$ where $\dot{\gamma}$ is the shear rate.

@@

$U$ and $V$ are the base streamwise (x) and wall normal (y) velocities
respectively. $T$ is the base stress tensor. We use Gauss-Labatto points in the
wall-normal ($y$) direction for the base profile, and decomposed the
disturbances to this flow into Fourier modes.  
\begin{equation}
    g(y) = \sum\limits_{n=-N}^{N} \widetilde{g}(y) e^{ikx + \lambda t}
\end{equation}
for all disturbance variables $g = u, v, p, \tau_{i,j}$. Where $N$ is the
number of Fourier modes, $k$ is the streamwise wavenumber of the disturbance
and $\lambda$ is the growth rate of the mode. This program uses two domains of
pseudo-spectral points to give increased resolution in the centre of the
simulation. The simulation was also repeated using a code for which the system
of points was stretched using: 
\begin{align}
    z_{u} &= \frac{y_{u}}{1-y_{u}} \\
    z_{b} &= \frac{y_{b}}{1+y_{b}} \\
    U     &= \tanh{z/\Delta} 
    \label{eq:KH_inf_profile}
\end{align}
This system ought to be completely insensitive to the boundary conditions on
the flow and so also behave as a free shear instability.

@@

Dependence of the instability on the channel walls persists at very small
$\Delta$. We believe that this is due to the inertial effects at low Reynold's
number being so weak. This means it is very easy to move the fluid even at
the shear rate present far from the centre of the channel. I think the
instability generated by the elasticity of the fluid needs the walls to dampen
it for some reason.

@@

\begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{KH_onset_beta_Wi}
    \caption{}
    \label{fig:KH_elastic_onset}
\end{subfigure}
~
\begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fin_dispersions_Wi50}
    \caption{}
    \label{fig:KH_finite_dispersions}
\end{subfigure}
\caption{
a) The change in the height of the dispersion relation as $\Wi$ is increased at
$\Rey = 0$ with $\Delta = 0.1$. This instability must be entirely elastic in
nature. b) The change in the height of the dispersion relation as the Reynolds
number is reduced at $\Wi=5$. At low $\Rey$ the instability is clearly still
present for large polymer concentrations, consistent with purely elastic
turbulence with $\Delta = 0.1$.  c) The value of the Weissenberg number at
which the velocity profile becomes unstable against $\beta$ at $\Delta = 0.1$.
The onset of the purely elastic instability moves to higher $Wi$ as $\beta$
increases. d) Dispersion relations at Wi =50 for $\Delta =0.1$.
}
\end{figure}

@@

In conclusion, we can say that there is a Kelvin-Helmholtz like instability for
purely elastic shear flows. The unstable region is much larger than that seen
in Newtonian fluids relative to the region of shear. The wavenumber of the
instability is much smaller in the viscoelastic case ($k<0.06$), suggesting a
larger minimal flow unit is required in viscoelastic fluids. This instability
is enhanced slightly by introducing a finite extensibility to the polymers but
remains of essentially the same form. The instability has grown to its maximum
by the time it reaches an effective Weissenberg number $ Wi \sim 50$ after
which it is saturated. The instability is also strongly dependent on the walls
at low effective Reynold's number.
